---
title: "Part 1: Processing of environmental data layers"
author: Genoveva Gonzalez-Mirelis
output: 
  html_notebook: 
    toc: yes
---

## Introduction

This part describes how to process the environmental layers to produce a predictor brick. The input files are the following: bathymetry, oceanographic variables, satellite image data, and landscape type. Note: Blending of oceanographic  layers (from NK800 and B800) is not included here, but feel free to contact me if you need to know the details of how that was done.

Another required input to run this Notebook is a shapefile of the area we are interested in modelling, which we call "mask".

All the required data is in the Data folder which you need to download and unzip directly in your working directory.

The output from this Notebook is an R stack of all layers, which is required as input for the spatial distribution model. This will be written to a new folder which will be created in this Notebook ("_PredictorStack") as an .RData file.

A bunch of intermediate files with extensions .RData, .tif, and .img will also be written in the PredictorData folder (at the root directory) which weren't there before.

## Setup and inputs

Specify data paths, load libraries.
```{r, message=FALSE, warning = FALSE}

setwd("/home/jovyan/project/")

#Libraries
library(RCurl)
library(rgdal)
library(raster)
library(rgrass7)
library(spatialEco) #vrm
library(cluster) #terrain classification
library(randomForest) #terrain classification


#download and unzip the Data folder

PredictorData = "Data/PredictorData"

```

Load and plot the modelling area.
```{r}

mask <- readOGR(file.path("Data", "BaseLayers","mask.shp"), verbose = FALSE)
plot(mask)

```

## Data processing

### Bathymetry and basic derivatives

Bathymetric data was downloaded from EMODnet Bathymetry (EMODnet Bathymetry Consortium (2018): EMODnet Digital Bathymetry (DTM). http://doi.org/10.12770/18ff0d48-b203-4a65-94a9-5fd8b0ec35f6). The tiles used were B5, B6, B7, B8, C5, C6, C7. These were mosaicked into one single layer elsewhere, and are provided as a single geotiff file in the data folder.

All variables that are computed at multiple scales are presented in Table 1.

In this block we derive basic terrain variables, including: topographic position index (at two scales), ruggosity, roughness, and slope.
```{r, message=FALSE, warning = FALSE}

bas_derivatives <- file.path(PredictorData, "t1.RData")

if(file.exists(bas_derivatives)) {
  print('basic terrain variables have already been derived')
  load(bas_derivatives)
  t1
  
  bathy <- raster(file.path(PredictorData,"Bathymetry", "BS_grid800m_utm33n.tif"))# load bathy file
  projection(bathy)<-"+proj=utm +north +zone=33 +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
  bathy <- mask(bathy,mask)
  
  writeRaster(bathy,file.path(PredictorData, "bathy.tif"), options="INTERLEAVE=BAND", overwrite=TRUE)
  
} else {

  #bathy
  ### AGGIUNTO:
  bathy <- raster(file.path(PredictorData,"Bathymetry", "BS_grid800m_utm33n.tif"))# load bathy file
  projection(bathy)<-"+proj=utm +north +zone=33 +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
  bathy <- mask(bathy,mask)
  ###
  ## derive terrain variables
  tpi <- terrain(bathy, opt=c('tpi')) # included in Table 1

  # TPI for different neighborhood size:
  tpiw <- function(x, w=5) {
	  m <- matrix(1/(w^2-1), nc=w, nr=w)
	  m[ceiling(0.5 * length(m))] <- 0
	  f <- focal(x, m)
	  x - f
  }

  tpi_b <- tpiw(bathy, w=15) # included in Table 1
  names(tpi_b) <- "tpi_b"
  
  tri <- terrain(bathy, opt=c('tri'))
  rough <- terrain(bathy, opt=c('roughness'))
  slope <- terrain(bathy, opt=c('slope'))

  t1 <- stack(slope,
              tpi,
              tpi_b,
              tri,
              rough
              )

  save(t1,file=file.path(PredictorData, "t1.RData"))
  
  rm(slope,
     tpi,
     tpi_b,
     tri,
     rough
     )
  
  t1
}
```
#writeRaster(t1,file.path(PredictorData, "t1.tif"), options="INTERLEAVE=BAND",     overwrite=TRUE)



### Additional terrain variables

In this block we derive multiscale terrain variables using GRASS, which you need to install prior ro running this part of the workflow. These are all included in Table 1

```{r message=TRUE, warning=TRUE}


method <- c("aspect",
              #"profc",
              #"planc",
              #"longc",
              #"crosc",
              #"minic",
              #"maxic",
              "feature") # only these two methods are used

win_param_scale <- c(3,19,33) # only these three window sizes are used

add_derivatives <- file.path(PredictorData, "t.RData")


try(system('grass74 -c EPSG:32633 /home/jovyan/grassdata/utm_wgs84_33N/'), silent = TRUE)
  
  
execGRASS('r.in.gdal', input=file.path(PredictorData,"Bathymetry", "BS_grid800m_utm33n.tif"),
            flags=c('overwrite','o','e'), output='elev')     # o= override the prjection parameters, and e=extend the location
  
execGRASS('g.region', raster='elev', flags="p", res=as.character(xres(bathy)))
  
  #win_param_scale <- c(1,3,5,7,9,11,13,15,17,19,21,23,25,27,29,31,33,35)
slope_tolerance <- 1.0
curvature_tolerance <- 0.0001
exponent <- 0.0
zscale <- 1.0
  
t <- stack()
  
for (j in 1:length(method)){
    
  for (i in 1:length(win_param_scale)){
      #print(j,i)
    execGRASS('r.param.scale', 
                input='elev', 
                output=paste(method[j],win_param_scale[i],sep="_"), 
                size=as.integer(win_param_scale[i]), 
                slope_tolerance=as.numeric(slope_tolerance), 
                curvature_tolerance=as.numeric(curvature_tolerance), 
                method=method[j], 
                exponent=as.numeric(exponent), 
                zscale=as.numeric(zscale), 
                flags=c('overwrite', 'quiet'))
      
      #p<-raster(readRAST(paste(method[j], win_param_scale[i],sep="_")))
      #t <- stack( t , p)
      
      print(paste(method[j],win_param_scale[i],sep="_"))
}}
    
execGRASS('i.group', group="stack", subgroup="stack",
          input=c("aspect_3","aspect_19","aspect_33",
                  "feature_3","feature_19","feature_33"))

execGRASS("r.out.gdal", input="stack",
          output=file.path(PredictorData, "stack.img"), format="HFA",
          flags=c('overwrite', 'quiet'))


t <- stack(file.path(PredictorData, "stack.img"))
names(t) <- paste(rep(method, each = length(win_param_scale)), win_param_scale, sep = "_")
projection(t)<-"+proj=utm +north +zone=33 +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"

save(t,file=file.path(PredictorData, "t.RData"))

t

#writeRaster(t,file.path(PredictorData, "t1.tif"), options="INTERLEAVE=BAND", overwrite=TRUE)
```

### Ruggedness

Implementation of the Sappington et al., (2007) vector ruggedness measure, as in the BTM tool.
```{r, message=FALSE}

rugged <- file.path(PredictorData, "vrm.RData")

if(file.exists(rugged)) {
  print('ruggedness has already been derived')
  load(rugged)
  vrm
  
} else {

  vrm <- vrm(bathy) 
  names(vrm) <- "vrm"
  
  save(vrm,file=file.path(PredictorData, "vrm.RData"))

  vrm

}

#writeRaster(vrm,file.path(output_PredictorData, "t1.tif"), options="INTERLEAVE=BAND", overwrite=TRUE)
```
